Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-02-28 paper/paper.tex
good draft

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Space for WHAM: a multi-region, multi-stock generalization of the
Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alexander C. Hansell}
\date{21 January, 2025}

@@

A state-space statistical approach and maximum marginal likelihood or
Bayesian fitting of stock assessment models allows estimation of time
and age-varying population attributes as random effects (Nielsen and
Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach
is considered an essential feature of gold-standard assessment models
that we use in tactical management of commercially important fish stocks
(Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and
Berg 2014) continues to be developed and remains widely used within the
International Council for the Exploration of the Sea (ICES) to assess
European fish stocks. Various state-space models are being used to
manage cod and plaice stocks in the waters of Eastern Canada (Perreault
et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole
Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is
now used to assess many fish stocks in the Northwest Atlantic Ocean
(NEFSC 2022a, 2022b; NEFSC 2024).

WHAM is an R package developed and maintained at NOAA's Northeast
Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
WHAM can be configured to fit a wide range of age-structured models from
traditional statistical catch-at-age models without any random effects
to models with several time and age varying process errors and effects
of environmental covariates on various population parameters. Like SAM,
WHAM models are built using the Template Model Builder package (TMB,
Kristensen et al. 2016) which provides a computationally efficient means
of fitting an extremely wide class of models with random effects. WHAM
has undergone active development since its creation and includes random
effects options for recruitment, inter-annual transitions of numbers at
age (hereafter referred to as ``survival''), fishery and index
selectivity, natural mortality, and catchability.

@@

We developed this extension of WHAM in concert with a research track
assessment process for black sea bass (\textit{Centropristis striata})
(NEFSC 2023). Previous black sea bass assessments assumed two stock
components (North and South), each assessed with a separate statistical
catch at age assessment. The stock was split into two regions to account
for spatial dynamics and improve model diagnostics (ASMFC 2016).
Additionally, black sea bass have complex spatial dynamics and make
seasonal migrations, moving inshore in the spring and offshore in the
fall (Moser and Shepherd 2009). Black sea bass distribution has been
linked to warming waters on the northwest shelf (Bell et al. 2015) and
it is hypothesized that black sea bass are especially susceptible to
climate change with likely increases in distribution and productivity
due to warming water temperatures (Hare 2016). Thus, a focus of the
research track assessment was to develop an assessment platform that
could capture multiple stock components, complex seasonal movement and
explore environmental drivers of population dynamics (NEFSC 2023).To
illustrate the usage of WHAM 2.0 we apply it to two stock components of
black sea bass off the coast of the NEUS using most of the model
assumptions that were accepted during the peer-review process. However,
here we evaluate evidence for alternative hypotheses of temporal
variation and effects specifically of bottom temperature on recruitment
and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions
further dimensionality is added to the class of possible equilbrium
reference points (e.g., Kapur et al. 2021). Short term projections are
also an important part of the management process and the separation of
observation and process errors in state-space models allows uncertainty
in projected attributes such as recruitment and spawning stock biomass
(SSB) without external simulation exercises. Finally, the inclusion of
explicit effects of environmental covariates in the model also requires
some assumptions on how they are treated in projections and reference
points. We demonstrate some of options of WHAM version 2.0 provide
biological reference points and short-term projections.

@@

Because individuals may be alive in one of several regions or harvested
in one of several fleets, it is helpful to consider these as distinct
categories or states and treat the number of individuals occurring in
each category over time as a multi-state model. Approaches to modeling
transitions among these states may treat time discretely (e.g., Arnason
1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
Commenges 1999; Andersen and Keiding 2002). Multi-state models can
define a probability transition matrix (PTM) that describes the
probability of individuals occurring in different states at the end of a
time interval \(\delta\), conditional on being in each of those state at
beginning of the interval. For fish populations, these states would be
defined as being alive in a particular region or deceased due to fishing
from a particular fleet or natural mortality. The time interval \(i\)
with duration \(\delta_i\) would be a season and the PTMs would be
uniquely defined for each stock by season \(i\), year \(y\), and age
\(a\) on January 1. Each row and column of the PTM correspond to one the
states: alive in region \(r\), dead in fleet \(f\), or dead from natural
causes. The probabilities in each row sum to unity and assume an
individual is in the corresponding state at the beginning of the
interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM
(\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation} where \begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
occurring in each region at the end of the interval, \begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
of being captured in each fleet during the interval, and
\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
dying due to natural mortality during the interval. We have the identity
matrix \(\mathbf{I}_{H}\) for the states for capture by each fleet and a
1 for the state for natural mortality because the probabilities of being
in one of the mortality states given starting the interval in that state
is unity (no zombies allowed).

WHAM uses these PTMs to model abundance proportions in each state rather
than true probabilities where numbers in each state would be multinomial
distributed as in a model for tagging data where fates of individual
fish are assumed independent. The PTMs determine the expected numbers 1)
in each state on January 1 of year \(t+1\) at age \(a+1\) given the
abundances at age \(a\) on January 1 of year \(t\), 2) captured over the
year in each fleet, 3) available to each index, and 4) alive at the time
and in the region where spawning occurs.

@@

When there is more than one region, WHAM can model survival and movement
as processes occurring sequentially or simultaneously. The sequential
assumption is used widely in spatially explicit model (e.g., Stock
Synthesis, Methot and Wetzel 2013). Under the sequential assumption,
survival and death occur over the interval and movement among regions
occurs instantly at either the beginning or the end of the interval.
WHAM is configured to have movement occur after survival and mortality:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*} where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\)
diagonal matrix of proportions surviving in each region (given they
start in that region) \begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*} and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\)
matrix of probabilities of moving from one region to another or staying
in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

Seasonality can be configured to accommodate characteristics of
spawning, movement, and fleet-specific behavior. The annual time step
can be divided into \(K\) (any number) seasons and the interval size
\(\delta_i\) for each season \(i\) does not need to be equal to any
other seasonal interval. Under the Markov assumption, the PTM of
surviving and moving and dying over \(K\) intervals
\(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the
product of the PTMs for each interval:
\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]

@@

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row
\(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for
each stock are the same as previous versions of WHAM (Miller and Stock
2020). If recruitment is assumed to be a function of SSB, it is only the
spawning population in region \(r_s\) at the time of spawning that
constitutes the SSB in the stock-recruit function. However, models can
configure spawning individuals to occur in other regions at the time of
spawning. Aside from treating recruitment as a random walk, the general
model for annual recruitment as random effects is \begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*} where\\
\[\text{SSB}_{s,y} =  \sum^A_{a=1}  w_{s,y,a} \text{mat}_{s,y,a} \mathbf{O}_{s,y,a,r_s}(t_s)' \mathbf{N}_{O,s,y,a}\]
where \(w_{s,y,a}\) is the mean weight at age of spawning individuals,
\(\text{mat}_{s,y,a}\) is the maturity at age, and
\(\mathbf{O}_{s,y,a,r_s}(t_s)\), the \(r_s\) column of the upper-left
submatrix of Eq. \ref{eq:ptm_spawn}, are the probabilities of surviving
and occurring in region \(r_s\) at time \(t_s\) given being alive in
each region at the start of the year.

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*} where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector
of region-specific errors for a given stock, which are independent
across stocks and regions, but the same correlation structures as
previous versions are possible across ages and years for a given stock
and region. When there is autocorrelation with age, WHAM now assumes
this applies only to ages \(a>1\) by default so that recruitment random
effects are independent of those for for the annual transitions of older
age classes. So the general covariance structure for a given stock at
ages \(a>1\) in region \(r\) is that of a two-dimensional first-order
autoregressive (2DAR1) process \begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*} and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since
recruitment for a given stock currently only occurs in one region
\(r_s\) there is only a single time-varying recruitment random effect
for each stock.

@@

The initial abundances at age can also be treated as independent or as
AR1 random effects. Defining the vector of initial abundance at age in
region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
where
\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement occur simultaneously, the
parameters are estimated with a log link function and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects
\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and(or)
region-to-region-specific and modeled most generally as 2DAR1 random
effects with age and(or) year where the covariance is \begin{equation*}
  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

>>>>

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\newcommand{\noop}[1]{}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Space for WHAM: a multi-region, multi-stock generalization of the
Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alexander C. Hansell}
\date{28 February, 2025}

@@

A state-space age-structured stock assessment models can be used to
estiamte time and age-varying population attributes as random effects
using maximum marginal likelihood or Bayesian fitting procedures
(Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This
estimation approach is considered an essential feature of gold-standard
assessment models that we use in tactical management of commercially
important fish stocks (Punt et al. 2020). The State-space Assessment
Model (SAM, Nielsen and Berg 2014) continues to be developed and remains
widely used within the International Council for the Exploration of the
Sea (ICES) to assess European fish stocks. Various state-space models
are being used to manage cod and plaice stocks in the waters of Eastern
Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the
Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and
Miller 2021) is now used to assess many fish stocks in the Northwest
Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

WHAM is an R package developed and maintained at NOAA's Northeast
Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
WHAM can be configured to fit a wide range of age-structured models from
traditional statistical catch-at-age models without any random effects
to models with several time and age varying process errors and effects
of environmental covariates on various population parameters. Like SAM,
WHAM models are built using the Template Model Builder package (TMB,
Kristensen et al. 2016) which provides a computationally efficient means
of fitting an extremely wide class of models with random effects. WHAM
has undergone active development since its creation and includes random
effects options for recruitment, inter-annual transitions of numbers at
age (hereafter referred to as ``apparent survival''), fishery and index
selectivity, natural mortality, and catchability.

@@

We developed this extension of WHAM in concert with a research track
assessment process for black sea bass (\textit{Centropristis striata})
(NEFSC 2023). Previous black sea bass assessments assumed two stock
components (North and South), each assessed with a separate statistical
catch at age assessment. The stock was split into two regions to account
for spatial dynamics and improve model diagnostics (ASMFC 2016).
Additionally, black sea bass have complex spatial dynamics and make
seasonal migrations, moving inshore in the spring and offshore in the
fall (Moser and Shepherd 2009). Black sea bass distribution has been
linked to warming waters on the northwest shelf (Bell et al. 2015) and
it is hypothesized that the distribution and productivity of the black
sea bass stock is especially susceptible to long-term changes in
temperature causing increases in abundance further north (Hare 2016).
Thus, a focus of the research track assessment was to develop an
assessment platform that could capture multiple stock components,
complex seasonal movement and explore environmental drivers of
population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we
apply it to two stock components of black sea bass off the coast of the
NEUS using most of the model assumptions that were accepted during the
peer-review process. However, here we evaluate evidence for alternative
hypotheses of temporal variation and effects specifically of bottom
temperature on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions
further dimensionality is added to the class of possible equilbrium
reference points (e.g., Kapur et al. 2021). Short term projections are
also an important part of the management process and the separation of
observation and process errors in state-space models allows uncertainty
in projected attributes such as recruitment and spawning stock biomass
(SSB) without external simulation exercises. Finally, the inclusion of
explicit effects of environmental covariates in the model also requires
some assumptions on how they are treated in projections and reference
points. We demonstrate some of options of WHAM version 2.0 to provide
biological reference points and short-term projections.

@@

Because individuals may be alive in one of several regions or harvested
in one of several fleets, it is helpful to consider these as distinct
categories or states and treat the number of individuals occurring in
each category over time as a multi-state model. Approaches to modeling
transitions among these states may treat time discretely (e.g., Arnason
1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
Commenges 1999; Andersen and Keiding 2002). We can define a probability
transition matrix (PTM) that describes the probability of individuals
occurring in different states at the end of a time interval \(\delta\)
(columns), conditional on being in each of those state at beginning of
the interval (rows). For fish populations, these states would be defined
as being alive in a particular region or deceased due to fishing from a
particular fleet or natural mortality. The time interval \(i\) with
duration \(\delta_i\) would be a season and the PTMs would be uniquely
defined for each stock by season \(i\), year \(y\), and age \(a\) on
January 1. Each row and column of the PTM correspond to one the states:
alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
The probabilities in each row sum to unity (must be in one of the
states) and assume an individual is in the state that corresponds to
that row at the beginning of the interval. Given \(n_R\) regions and
\(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a
function of sub-matrices is \begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation} where \begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
occurring in each region at the end of the interval, \begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
of being captured in each fleet during the interval, and
\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
dying due to natural mortality during the interval. We have the identity
matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a
1 for the natural mortality state because the probabilities of being in
one of the mortality states given starting the interval in that state is
unity (no zombies allowed).

WHAM uses these PTMs to model abundance proportions in each state rather
than true probabilities where numbers in each state would be multinomial
distributed as in a model for tagging data where fates of individual
tagged fish are assumed independent. The PTMs determine the expected
numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\)
given the abundances at age \(a\) on January 1 of year \(t\), 2)
captured over the year in each fleet, 3) available to each index, and 4)
alive at the time and in the region where spawning occurs.

@@

When there is more than one region, WHAM can model survival and movement
as processes occurring sequentially or simultaneously. The sequential
assumption is used widely in spatially explicit model (e.g., Stock
Synthesis, Methot and Wetzel 2013). Under the sequential assumption,
survival and death occur over the interval and movement among regions
occurs instantly at either the beginning or the end of the interval.
WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*} where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\)
diagonal matrix of proportions surviving in each region (given they
start in that region) \begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*} and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\)
matrix of probabilities of moving from one region to another or staying
in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

Seasonality can be configured to accommodate characteristics of
spawning, movement, and fleet-specific behavior. The annual time step
can be divided into \(K\) seasons and the interval size \(\delta_i\) for
each season \(i\) does not need to be equal to any other seasonal
interval. Under the Markov assumption, the PTM of surviving and moving
and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the
entire year) is just the product of the PTMs for each interval:
\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]

@@

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row
\(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for
each stock are the same as previous versions of WHAM (Miller and Stock
2020). If recruitment is assumed to be a function of SSB, it is only the
spawning population in region \(r_s\) at the time of spawning that
constitutes the SSB in the stock-recruit function. However, models can
configure spawning individuals to occur in other regions at the time of
spawning. Aside from treating recruitment as a random walk, the general
model for annual recruitment as random effects is \begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*} where\\
\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
and \(w_{y,a}\) is the mean weight at age of spawning individuals,
\(\text{mat}_{y,a}\) is the maturity at age, and
\(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and
occurring in region \(r_s\) at time \(t_s\) given being alive in each
region at the start of the year (the \(r_s\) column of the upper-left
submatrix of Eq. \ref{eq:ptm_spawn}).

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*} where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector
of region-specific errors for a given stock, which are independent
across stocks and regions, but the same correlation structures as
previous versions are possible across ages and years for a given stock
and region. When there is autocorrelation with age, WHAM now assumes
this applies only to ages \(a>1\) by default so that recruitment random
effects are independent of those for for the annual transitions of older
age classes. So the general covariance structure for a given stock at
ages \(a>1\) in region \(r\) is that of a two-dimensional first-order
autoregressive (2DAR1) process \begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*} and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since
recruitment for a given stock currently only occurs in one region
\(r_s\) there is only a single vector of time-varying recruitment random
effects for each stock.

@@

The initial abundances at age can also be treated as independent or AR1
random effects. Defining the vector of initial abundance at age in
region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
where
\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement occur simultaneously, the
parameters are estimated with a log link function and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of an intercept
or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and
environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects
\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and
region-to-region-specific and modeled most generally as 2DAR1 random
effects with age and(or) year where the covariance is \begin{equation*}
  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

There is currently no likelihood component for tagging data. Therefore,
movement parameters would usually either need to be fixed or assumed to
have some prior distribution, possibly based on external parameter
estimates. WHAM includes prior distributions for the season and
region-to-region specific (mean) movement parameters and treats them as
random effects with the mean defined by the initial value of the fixed
effect counterpart and standard deviation \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*} When priors are used, the movement is defined instead
as \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}
